Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela

### 2024-11-21 agujournaltemplate.tex
first draft of methods section

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{The CARIACO Time Series}
  %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time-Series Program was established to investigate the link between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large ($\sim$ 160 km long, 70 km wide) and deep ($\sim$ 1,400 m) basin, located on the north-eastern Venezuelan continental shelf, bound to the north by a sill at a mean depth of about 100 m. It is a relatively productive area ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes that are higher than BATS and HOT locations, leading to the accumulation of anoxic bottom waters below 200 m depth.

    \subsubsection{Niskin Bottle Dataset}
        Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W).

    \subsubsection{CTD Dataset}

    \subsubsection{Phytoplankton Microscopy}

    \subsubsection{ERA 5 climate data for wind}

    \subsubsection{AMO and MEI v.2 Climate indices}

        %% METHODS
    \subsubsection{Z-Scores calculation}

    \subsubsection{Clustering and NMDS}

        \subsubsection{Gradient Forest}

%
\section{Data} (Or section title might be a descriptive heading about data)
%
% \section{Results} (Or section title might be a descriptive heading about the
% results)
%

\section{Draft Results}
%Text here ===>>>
Below here I will put the draft plots to give an overview over the current state of the results. As of now, all plots are in a draft state, sometimes missing appropriate units. Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.

>>>>

        %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time-Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (about 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at 8 standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela), and from mid-1997 to the present, by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. The data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ data showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged. For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.

    For phytoplankton species identification by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.

    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).

    From the CARIACO Time Series data, we used only those variables, that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.4°N,64.8°W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)

    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all depth discrete measurements, the data was interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated and then grouped by Genus via the sum of interpolated counts.

    \subsection{Data analysis}    
    To give a general overview of the data and trends over the time series, we calculated yearly means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). The cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the dataset more homogeneous and more robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}. Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\'s evenness $J' = -\sum p_i ln( p_i )/ln(H')$ based on the genera present for each interpolated and integrated monthly sampling.

    %RUE/BDEF ANALYSIS
    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\'s Index, each integrated at the four depth intervals across each monthly sampling. $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
    %here I want to extend to glmm that correct for depth and seasonality for the final results

    \subsubsection{Clustering and NMDS}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of yearly aggregates of genera observed integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify ranges of the predictors where significant changes occur in indicator variables, which signal a threshold for community turnover in the phytoplankton community. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. These analyses were conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, -u 10m wind component, the 21°C Isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which Gradient Forest methodology takes into account by implementing a conditional permutation of correlated predictors \cite{ellis_gradient_2012}. To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run. % see figure in supplementary with initial runs TODO

\section{Draft Results}

%from Mutshinda et al. 2013
Phytoplankton species richness varied from 4 to 100 species per cruise, with a mean of 40 over the CARIACO time series (167 cruises from November 1995 to March 2011). Over a third of the species (187 of 525) were only observed during one or two cruises in the entire study.

%Text here ===>>>
Below here I will put the draft plots to give an overview over the current state of the results. As of now, all plots are in a draft state, sometimes missing appropriate units. Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.
